A comparison on R tools for nonlinear least squares modeling

### 2023-03-02 RefactoringNLS/ComparingNLStools.Rmd
new submission

>>>>

## Scope of our comparison

The tools we will compare are the base-R `nls()` function 
and a number of packages that are available on the CRAN repository.

Important:

nlsr
minpack.lm
gslnls
nls2

Worth commentary:

crsnls
NISTnls
nlshelper
nlsic
nlsMicrobio
nlsmsn
nls.multstart
nlstac
nlstools
onls

Mention only:

easynls

## Comparison dimensions

**nlsr** has functions to find a set of parameters that minimizes the sum of 
squares (or deviance) 
of differences between a model specified either as a formula using function 
`nlxb()` or as both residual
and Jacobian R functions using function `nlfb()`. As described below, the 
Jacobian may be approximated
by particular control settings.

Existing R tools that are similar to `nlxb()` are `nls()` and `minpack.lm::nlsLM()`;  
`minpack.lm::nls.lm()` is similar to `nlfb()`.

There are several important differences in approach and results, some of which 
are detailed in this document.

### Output object

The output object of `nlxb()` is smaller than that of `nls()` and is focused on 
the solution and its properties
rather than the **modeling** task. That is, package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. `nls()` and `nlsLM` return an object of 
class `nls`, which allows for
a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

### Jacobian calculation

Internally, `nlxb()` uses symbolic and automatic differentiation tools to create 
the residual and Jacobian functions needed for `nlfb()` then uses it to solve 
the relevant nonlinear least squares minimization problem.

As with other nonlinear least squares packages, we provide the Jacobian code as 
the "gradient" attribute of the Jacobian function. This lets us embed the code 
for the Jacobian as this attribute of the **residual** function so that the call 
to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

Generally `nls()` and `nlsLM()` use a fairly simple forward difference 
approximation of derivatives for the Jacobian, though a central approximation 
can be specified in control parameters. Package `nlsr` provides four approximation 
options, with a further control `ndstep` for the size of the
step used in the approximation.

### Stabilization of Gauss-Newton computations

The iteration in all the major tools mentioned is the solution of the Gauss-Newton 
equations. `nls()` uses a variant of an approach suggested by @Hartley61, while 
`nlsr` and `minpack.lm` use variants of the method published by @Marquardt1963, 
though it was suggested earlier by @Levenberg1944. The details of the method in 
`nlsr` are discussed in  @JNABRefactoringNLS22.
Though it is unlikely to be useful in general, control settings for `nlxb()` or `nlfb()`
allow for those routines to perform a variety of Hartley and Marquardt algorithms. This
has been useful for learning about the algorithms, but tangential to simply finding
parameters of nonlinear models.

In general, the Levenberg-Marquardt stabilization is important in obtaining solutions.
The default method of `nls()` terminates with `singular gradient` unnecessarily in 
too many instances.

### Programming language

An important choice made in developing `nlsr` was to code entirely within the R 
programming language. `nls()` uses a mix of R, C and Fortran, as does `minpack.lm`. 
Generally, I believe that keeping to a single programming language allows for 
easier maintenance and upgrades. On the other hand, R is usually somewhat slower 
in performing computations because it keeps track of names and structures and 
because it is usually interpreted rather than compiled. In recent years the 
performance penalty for using code entirely in R has been much reduced
with the just-in-time compiler and other improvements, so that using an all-R 
package offers acceptable performance. Indeed, in `nlsr` the use of R may be 
less of a performance cost than the attempt to ensure the solution has been found, 
which forces more iterations to be used.

---OLD VERSION BELOW ------------------------
